Geometric variational problems

### 2020-11-09 NSF2020/Description/specificaim3_ver2.tex

\newpage

@@

\begin{wrapfigure}[21]{r}{3.0in}
%\vspace*{+15pt}
\centerline{\includegraphics[width=3.0in]{figures/integrands}}
\vspace*{-10pt}
\caption{{\footnotesize The Fourier modes of $K(x_0,y)$ and $K_2(x_0,y)$
  where $\|x_0\| = 0.99$ and $y \in \partial\Omega$, where
  $\partial\Omega$ is the unit circle. The trapezoid rule applied to the
  red integrand has large amounts of error, but the trapezoid rule is
  much more accurate when applied to the blue integrand. Since the
  difference between the red and blue curves can be accurately computed
  with the barycentric quadrature rule, the quadrature error applied to
  layer potentials for the screen Laplace equation will be uniformly
  bounded in $\Omega$.}}
\label{fig:integrands}
\end{wrapfigure}
Nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and this results in nearly-singular integrands. Quadrature
methods for these integrands has received a lot of attention in
two and three dimensions~\cite{alpert, kapur, sidi, duffy, bruno1,
bruno2, davis_1984, graglia_2008, hackbusch_sauter_1994, jarvenpaa_2003,
khayat_2005, schwab_1992, ying_2006, beale1, beale2, goodman_1990,
haroldson_1998, lowengrub_1993, schwab_1992, ggq1, ggq2, ggq3,
helsing_2008a, helsing_integral_2009, helsing_tutorial_2012,
klockner2013jcp, qbx2, wala2019jcp, af2018sisc, siegel2018jcp,
rachh2017jcp, ding2019arxiv, bar2014}. The trapezoid rule is the
workhorse for two-dimensional BIEs since it achieves spectral accuracy
when integrands are not nearly-singular~\cite{tre-wei2014}. To address
nearly-singular integrands of two-dimensional BIEs, we will use a {\em
barycentric quadrature rule} that only requires a slight modification of
the trapezoid rule~\cite{ioa-pap-per1991}. This method requires the
layer potential satisfy Laplace or Stokes
equations~\cite{bar-wu-vee2015, chi-moo-qua2020}. The PIs will extend
this quadrature method to layer potentials for the screened Laplace
equation. This will be done by recognizing that the fundamental solution
of the screened Laplace equation can be decomposed as $K(x,y) = K_1(x,y)
+ K_2(x,y)$, where $K_1(x,y) = -\log\|x - y\|$ and $K_2(x,y) = K(x,y) +
\log\|x - y\|$. Using this decomposition, layer potentials involving
$K_1(x,y)$ can be accurately computed with the barycentric quadrature
rule~\cite{ioa-pap-per1991}, and layer potentials involving $K_2(x,y)$
can be accurately computed with the trapezoid rule since the kernel is
bounded for all $x$ and $y$. The proposed method is demonstrated in
Figure~\ref{fig:integrands} where the amplitude of the Fourier modes of
$K(x_0,y)$ and $K_2(x_0,y)$ are plotted. The quadrature of the trapezoid
rule is the sum of the Fourier amplitudes not captured at the
resolution.

@@

A numerical issue when simulating the self-assembly of amphiphilic
particles in a viscous solvent is the collision between the
particles. The hydrophobic attraction potential drives the amphiphilic
particles towards one another so to minimize exposure to the solvent,
and this leads to physical contact in finite time. Such particle
collisions in a dense rigid body suspension is a great challenge and can
be a bottleneck in large-scale simulations. We propose two algorithms to
remedy this computational challenge: high-order adaptive time stepping
and repulsion forces.

PI Quaife developed a high-order adaptive time stepping method for
hydrodynamic suspensions~\cite{qua-bir2016} and it has served as a
reliable method to simulate processes including mixing and adhesion in
suspensions~\cite{qua-vee-you2019, kab-qua-bir2017}. The method uses
a single-step high-order time stepping method and a computationally
cheap estimate of the error. The proposed work will use a spectral
deferred correction method~\cite{dut-gre-rok2000} since it can
iteratively apply a low-order method to achieve high-order accuracy. To
estimate the error, we will compute the total force and torque of the
system which is computationally cheap and physically zero. The PIs
experience is that error estimates based on physical constraints such as
force- and torque-free conditions appropriately adjust the time step
size so that the dynamics are adequately resolved without the
computational expense of strategies such as embedded Runge-Kutta
methods, step-doubling, and Richardson extrapolation.

@@

% ----------------------------------------------------------------------
\subsubsection{Fluctuating hydrodynamics}
\label{subsec:fluctuating}
Once algorithms that address quadrature, adaptive time stepping, and
contact are implemented, we will embark on incorporating fluctuating
hydrodynamics~\cite{Bao17,Bao18} which are important even at scales of
the some amphiphilic particles.

\todo[inline]{Using DLP formulation for fluctuating hydrodynamics so
that we have a second-kind BIE}
%\textcolor{red}{
%Once the collision-resolution algorithm is implemented, we will check
%the computing performance first before we embark on incorporating
%fluctuating hydrodynamics~\cite{Bao17,Bao18}, which might be important
%even at the scales of the amphiphilic particle size in some cases. We
%propose to extend the high-order time-integration scheme for a
%stochastic differential equation~\cite{fu2015pre} to this system.
%}

>>>>

%\newpage

@@

\begin{wrapfigure}[21]{r}{3.0in}
%\vspace*{+15pt}
\centerline{\includegraphics[width=3.0in]{figures/integrands}}
\vspace*{-13pt}
\caption{{\footnotesize The Fourier modes of $K(x_0,y)$ and $K_2(x_0,y)$
  where $\|x_0\| = 0.99$ and $y \in \partial\Omega$, where
  $\partial\Omega$ is the unit circle. The trapezoid rule applied to the
  red integrand has large amounts of error, but the trapezoid rule is
  much more accurate when applied to the blue integrand. Since the
  difference between the red and blue curves can be accurately computed
  with the barycentric quadrature rule, the quadrature error applied to
  layer potentials for the screen Laplace equation will be uniformly
  bounded in $\Omega$.}}
\label{fig:integrands}
\end{wrapfigure}
Nearly touching bodies is ubiquitous in self-assembly of amphiphilic
particles, and this results in nearly-singular integrands. Quadrature
methods for these integrands has received a lot of attention in
two and three dimensions~\cite{alpert, kapur, sidi, duffy, bruno1,
bruno2, davis_1984, graglia_2008, hackbusch_sauter_1994, jarvenpaa_2003,
khayat_2005, schwab_1992, ying_2006, beale1, beale2, goodman_1990,
haroldson_1998, lowengrub_1993, schwab_1992, ggq1, ggq2, ggq3,
helsing_2008a, helsing_integral_2009, helsing_tutorial_2012,
klockner2013jcp, qbx2, wala2019jcp, af2018sisc, siegel2018jcp,
rachh2017jcp, ding2019arxiv, bar2014}. The trapezoid rule is the
workhorse for two-dimensional BIEs since it achieves spectral accuracy
when integrands are not nearly-singular~\cite{tre-wei2014}. To address
nearly-singular integrands of two-dimensional BIEs, we will use a {\em
barycentric quadrature rule} that only requires a slight modification of
the trapezoid rule~\cite{ioa-pap-per1991}. This method requires the
layer potential satisfy Laplace or Stokes
equations~\cite{bar-wu-vee2015, chi-moo-qua2020}. The PIs will extend
this quadrature method to layer potentials for the screened Laplace
equation. This will be done by recognizing that the fundamental solution
of the screened Laplace equation can be decomposed as $K(x,y) = K_1(x,y)
+ K_2(x,y)$, where $K_1(x,y) = -\log\|x - y\|$ and $K_2(x,y) = K(x,y) +
\log\|x - y\|$. Using this decomposition, layer potentials involving
$K_1(x,y)$ can be accurately computed with the barycentric quadrature
rule~\cite{ioa-pap-per1991}, and layer potentials involving $K_2(x,y)$
can be accurately computed with the trapezoid rule since the kernel is
bounded for all $x$ and $y$. The proposed method is demonstrated in
Figure~\ref{fig:integrands} where the amplitude of the Fourier modes of
$K(x_0,y)$ and $K_2(x_0,y)$ are plotted. The quadrature of the trapezoid
rule is the sum of the Fourier amplitudes not captured at the
resolution.

@@

A numerical issue when simulating the self-assembly of amphiphilic
particles is avoiding particle collision. The hydrophobic attraction
potential drives the amphiphilic particles towards one another so to
minimize exposure to the solvent, and this leads to physical contact in
finite time. Such particle collisions in a dense rigid body suspension
is a great challenge and can be a bottleneck in large-scale simulations.
We propose two algorithms to remedy this computational challenge:
high-order adaptive time stepping and repulsion forces.

PI Quaife developed a high-order adaptive time stepping method for
hydrodynamic suspensions~\cite{qua-bir2016} and it has served as a
robust method to simulate processes including mixing and adhesion in
suspensions~\cite{qua-vee-you2019, kab-qua-bir2017}. The method uses
a single-step high-order time stepping method and a computationally
cheap estimate of the error. The proposed work will use a spectral
deferred correction method~\cite{dut-gre-rok2000} since it iteratively
applies a low-order single-step method to achieve high-order accuracy.
To estimate the error, at each time step we will compute the total force
and torque of the system which is computationally cheap and physically
zero. The PIs' experience is that error estimates based on physical
constraints such as force- and torque-free conditions appropriately
adjust the time step size so that the dynamics are resolved without the
computational expense of techniques such as embedded Runge-Kutta
methods, step-doubling, and Richardson extrapolation.

@@

% ----------------------------------------------------------------------
\subsubsection{Fluctuating hydrodynamics}
\label{subsec:fluctuating}
Once algorithms that address quadrature, adaptive time stepping, and
contact are implemented, we will embark on incorporating fluctuating
hydrodynamics that are important even at scales of the some amphiphilic
particles. Bao et al.~\cite{Bao17,Bao18} developed two-dimensional BIE
formulations for Brownian rigid bodies. To satisfy the covariance of the
rigid body velocities, an ill-conditioned first-kind integral equation
method is used. The proposed work will consider a double-layer potential
formulation that has much better conditioning properties. This strategy
was attempted in~\cite{Bao18}, but they claim that a necessary
regularization of the covariance matrix leads to a drastic loss of
accuracy in numerical fluctuation-dissipation balance. However, they
only one regularization strategy was considered, and as PI Quaife has
shown in previous work, the regularization choice significantly affects
the resulting physics~\cite{ong-chr-qua2017}. Therefore, different
regularization strategies will be investigated with the goal of
controlling the accuracy of the fluctuation-dissipation balance.
